---
title: "Assignment 1"
author: "79546 - Stephen K. Ng'etich"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---


```{r echo=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6) 

```


# Pre-requisite {-}
## Load packages
```{r message = FALSE,warning = FALSE}
# Clear variables
rm(list=ls())
library(markovchain)
library(diagram)
library(castor)

```
\newpage

# Q1  

## Generate the transition probabilities
the Markov chain state space $S ={H,D,Y}$ where H represents Harvard, D for Dartmouth and y for Yale
```{r}
states = c("H", "D", "Y")
trans_mat1 <- matrix(c(.6, 0, .4,
              .3, .5, .2,
              .35, .35, .3), nrow = 3, byrow = TRUE)

rownames(trans_mat1) = states
colnames(trans_mat1) = states
print(trans_mat1)

```
## Draw the transition states with their respective probabilities

```{r}
mk_chain1 = new("markovchain",
                 states=states,
                 transitionMatrix=trans_mat1,
                 name="HDY",
                )

plotmat(trans_mat1,relsize = 0.75)
```

## Find the probability that the grandson of a man from Harvard went to Harvard   

```{r}
trans_mat2 = trans_mat1 %*% trans_mat1
trans_mat2

```
The probability that the grandson of a man from havard went to have is `0.5`

## Modify the above by assuming that the son of a Harvard man always went to Harvard.  
```{r}
states = c("H", "D", "Y")
trans_mat3 <- matrix(c(1, 0, 0,
              .3, .5, .2,
              .35, .35, .3), nrow = 3, byrow = TRUE)

rownames(trans_mat3) = states
colnames(trans_mat3) = states
print(trans_mat3)
```
## Determine the steady-state probabilities
```{r}
steadyStates(mk_chain1)
```
# Q2  
## Set up the matrix of the transistion probabilities  
```{r}
states1 = c("professional", "skilled labourer", "unskilled labourer")
trans_mat3 <- matrix(c(.7, .25, .05,
                      .22, .58, .2,
                      .15, .15, .7), nrow = 3, byrow = TRUE)

rownames(trans_mat3) = states1
colnames(trans_mat3) = states1
print(trans_mat3)
```
## Draw the transition states with their respective probabilities
```{r}
plotmat(trans_mat3,relsize = .75,box.cex = 0.54,cex = 0.7)
```
## Find the probability that a randomly chosen grandson of an unskilled labourer is a professional man.  

```{r}
trans_mat4 = trans_mat3 %*% trans_mat3
print(trans_mat4)
```
The probability is `0.243`  

## Determine the steady-state probabilities  
```{r}

mk_chain2 = new("markovchain",
                 states=states1,
                 transitionMatrix=trans_mat4,
                 name="Labourer",
                )
steadyStates(mk_chain2)
```
# Q3  

## Is this chain irreducible?    

Yes, since one chain can get from one state to any other state in the chain

## Is this chain aperiodic?  

Yes, the greatest common divisor when moving from state 1 and back is one

## Find the stationary distribution for this chain.  
```{r}

states3 = c("1", "2", "3")
trans_mat5 <- matrix(c(0.5, 0.25, .25,
                      1/3, 0, 2/3,
                      0.5, 0.5, 0), nrow = 3, byrow = TRUE)

rownames(trans_mat5) = states3
colnames(trans_mat5) = states3


stationary_states = get_stationary_distribution(trans_mat5)
print(stationary_states)


mk_chain3 = new("markovchain",
                 states=states3,
               transitionMatrix=trans_mat5,
                 name="",
              )

# test correctness (steady states* trasnsition matrix should be 0, apart from rounding errors)
cat(sprintf("max(abs(steady states* trasnsition matrix)) = %g\n",max(abs(steadyStates(mk_chain3) %*% trans_mat5))))

#Check if it is a limiting distribution
apply(steadyStates(mk_chain3),1,sum)

steadyStates(mk_chain3)

```

## Is the stationary distribution a limiting distribution for the chain?  

Yes, since the chain is irreducible and aperiodic

# Q4  
## Stock market price data  
Yes, Stock Market price is highly volatile 

## Appraisal of a Secondary school mathematics teacher      
Yes, The outcome of the appraisal is dependent on past performances 

## Collaborative filtering on a database of movie reviews: for example, Netflix challenge: predict about how much someone is going to enjoy a movie based on their and other users’ movie preferences  
No, User rarely change their movie preferences  

## Daily weather forecast in Nairobi   
Yes, the daily weather  

## Optical character recognition   
Yes, combination of characters have different generate different meanings.  

## Cost of gemstones in Bangladesh    
Yes, cost of gemstone change over time due to demand on the local market and international market.  

# Q5  

$\pi$ and $A$ updated based on the following update rules
$$\pi_k = \frac{\gamma(z_{1k})}{\sum^K_{j=1}\gamma(Z_{1j})}$$
$$A_{jk} = \frac{\sum^K_{j=1}\xi(z_{t-1},_jz_{tk})}{\sum^K_{m=1}\sum^K_{t=2}\xi(z_{t-1},_jz_{tm})}$$
given $$\xi(z_{t-1},_jz_{t}) == \frac{\alpha(z_{t-1})P(x_t|z_t)P(z_t|z_{t-1}\beta(z_t))}{P(X)}$$
if  $A_{jk}$ is 0, then $\xi(z_{t-1},_jz_{t})$ is also 0, which makes the subsequent updates in the EM algorithm 0















