---
title: "Assignment 1"
author: "79546 - Stephen K. Ng'etich"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---


```{r echo=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6) 
```


# Pre-requisite
## Load packages
```{r message = FALSE,warning = FALSE}
# Clear variables
rm(list=ls())
library(markovchain)
library(diagram)
```


# Q1  

## Generate the transition probabilities
the Markov chain state space $S ={H,D,Y}$ where H represents Harvard, D for Dartmouth and y for Yale
```{r}
states = c("H", "D", "Y")
trans_mat1 <- matrix(c(.6, 0, .4,
              .3, .5, .2,
              .35, .35, .3), nrow = 3, byrow = TRUE)

rownames(trans_mat1) = states
colnames(trans_mat1) = states
print(trans_mat1)
```
## Draw the transition states with their respective probabilities

```{r}
mk_chain1 = new("markovchain",
                 states=states,
                 transitionMatrix=trans_mat1,
                 name="HDY",
                )

plotmat(trans_mat1,relsize = 0.75)
```

# Q4  
## Stock market price data  
Yes, Stock Market price is highly volatile 

## Appraisal of a Secondary school mathematics teacher      
No, 

## Collaborative filtering on a database of movie reviews: for example, Netflix challenge: predict about how much someone is going to enjoy a movie based on their and other users’ movie preferences  
No, User rarely change their movie preferences  

## Daily weather forecast in Nairobi  
Yes, the weather forecast of the a given day is related to the previous day
## Optical character recognition 
Yes, when characters are observed in sequence they form words which have different meaning depending on the order.
## Cost of gemstones in Bangladesh  
No, 
