---
title: "CAT 1"
author: "79546 Stephen Ng'etich"
output:
  pdf_document:
    keep_tex: yes
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xspace}
---
# Q1    
## a) Peter bets 5 dollars each time (timid strategy). 

$$p=0.55,q=0.45,N=8,z=5$$  

```{r}
p = 0.55
q = 0.45
N = 8
i = 5

timid_results = (1-(q/p)^i) / (1-(q/p)^N)

timid_results
```
## b) Peter uses a bold strategy
Since peter uses a bold strategy i.e play with the money earned in each round.in this case he will play  4 round to get to 8 dollars.
```{r}
q_0 <- 0
q_8 <- 1

p <- 0.55
q <- 0.45

q_4 <- p*(q_8)+q*(q_0)
q_2 <- p*(q_4)+q*(q_0)
result_bold <- p*(q_2)+q*(q_0)

result_bold

```
## c) Which strategy is better timid or bold

Timid Strategy is better than the bold strategy



# Q2    

The gambler's ruin problem:
player A starts with fortune $j$ dollars and bets 1 dollar until he either loses all the fortune or reach a $N$ dollars fortune and then quit.

let $A_j$ represent the the event that player A win starting with j dollars.
$$ x_j = P(A_j) = \text{probability to win starting with a j} = \text{probability to reach N before reaching 0 starting from j}$$

Using conditional probability and condition on what happens at the first game, win,lose or tie,every game will have $$P(win)=p, P(lose)=q,P(tie)=r$$ which yields $$\begin{aligned} x_j & = P(A_j) \\ & = P(A_j|win)P(win) + P(A_j|lose)P(lose)+P(A_j|tie)P(tie) \\ & =  x_j \times p + x_j \times q + x_j \times r \end{aligned}$$
note that $x_0 = P(A_0) = 0$ since their nothing more to gamble and $x_N = P(A_N)$ since player A has reached their goal and then stop playing hence the equation $p+q+r=1$   
Gamblers Ruin
$$px_{j+1} - (p+q)x_j+qx_{j-1} = 0, x(0)=0,x(N)=1$$ 

.Rewriting the equation in second order equation with $x_j = \alpha^j$ we find the quadratic equation $$p\alpha^2 - (p+q)\alpha+q=0$$ with solutions $$\alpha = \frac{p+q \pm \sqrt{(p+q)^2-4pq}}{2p} = \frac{p+q \pm \sqrt{p^2+q^2-4pq}}{2p} = \frac{p+q \pm\sqrt{(p+q)^2}}{2p} = \bigg\{ \begin{array}{ll}1 \\ q/p\end{array} $$
if $p \ne q$ there arer two solutions and so the general is given by $$x_n = C_1 1^n + C_2 \bigg(\frac{q}{p} \bigg)^n$$. Using $X_o=0 \text{ (lose) },X_N=1 \text{ (win) }$ to determine the constants $C_1$ and $C_2$.
$$0=C_1+C_2,  and 1=C_1+C_2\bigg(\frac{q}{p}\bigg)^N$$ which gives $$C_1 = -C_2 = \bigg(1-\bigg(\frac{q}{p}\bigg)\bigg)^{-1}$$
hence $$ \text{Gambler's ruin probabilities: } x_n = \frac{1-(q/p)^n}{1-(p/q)^N} \, \, p\ne q$$

if the game is fair i.e. $p=q$ the gambler ruin probabilities $X_j$ simplify to $$x(j+1)-2x(j)+x(j-1)$$ which gieve the quadratic equation $$\alpha^2 - 2\alpha +1$$ with the only root $\alpha =1$.The genral solution is $$x(j)=C_1+C_2j$$ with $x(0) = 0 $ and $x(N) =1$ which results $$ \text{Gambler's ruin probabilities: } x_n = \frac{j}{N} \, \, if \,\,p = q$$

in summary the game is fair 
$$
\begin{aligned} 
\text{fair  if } p=q \\ 
\text{subfair if  } p < q \\ 
\text{superfair if } p > q 
\end{aligned}
$$

# Q3     

Tom starts with $5, and p = 0.63: What is the probability that Tom obtains a fortune
of N = 12 without going broke?

$$i=5,N=12 \text{ and } q=1-p=0.37$$
hence
$$\frac{q}{p}=\frac{37}{63}$$
$$P_2 = \frac{1-(37/63)^5}{1-(37/63)^{12}} = \frac{0.93012}{0.99832} = 0.9317$$
What is the probability that Tom will become infinitely rich?

$$1-(q/p)^i = 1-(37/63)^5 = 0.93012 $$

If Tom instead started with i = $2, what is the probability that he would go broke?

$$ \text{The probability he becomes rich is } 1-(q/p)^i  = 1-(37/63)^2= 0.6550$$

# Q4  

The probability that the stock goes up by 15 before going down by 6.computing $p(a)$
$$p=0.6,q=1-p=0.4,a=15,b=6$$
$$p(a)=\frac{1-(\frac{q}{p})^b}{1-(\frac{q}{p})^{a+b}} = \frac{1-(\frac{0.4}{0.6})^6}{1-(\frac{0.4}{0.6})^{15+6}} = \frac{0.9122}{0.9998} =0.9914 $$
